Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-20 ind_files/06_nontechnical_discussion.tex
Minor rephrasing

At first thought, the idea of an evolutionary force that selects individuals with lower birth and death rates over individuals with higher birth and death rates may be reminiscent of notions in life-history evolution like $r$ vs $K$ selection or selection on the pace of life~\citep{stearns_evolution_1977}. However, it is not possible to conclude whether this similarity reflects some deep principle or whether it is just superficial based solely on the work conducted in this thesis. Models in life-history theory are often primarily concerned with spatiotemporally fluctuating external environments, and thus the stochasticity at the core of life-history models is extrinsic to the population. Ecological frameworks such as modern coexistence theory, which also deal with questions about similar population dynamics and would benefit from a first principles stochastic birth-death formulation, also typically work with fluctuating external environments~\citep{chesson_stabilizing_1982,chesson_multispecies_1994,barabas_chessons_2018, johnson_resolving_2022}. I have entirely neglected such extrinsic factors in my formalism. In principle, it is possible to make the birth and death rates \eqref{nD_functional_forms_for_replicator} in my framework also depend on a temporally varying external environment $\mathbf{E}(t)$ (whose variation may possibly depend on the population $\mathbf{n}(t)$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. If the variation of the environment $\mathbf{E}(t)$ has some associated stochasticity, the complete dynamics of the system would be the result of interactions between two qualitatively different forms of noise --- \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population --- and can consequently be quite complex and intricate, as shown in both theoretical~\citep{gokhale_eco-evolutionary_2016,wienand_evolution_2017} and empirical\citep{chavhan_larger_2020, chavhan_interplay_2021} studies. Thus, while integrating the birth-death framework I outline here with ecological ideas such as the pace-of-life syndrome~\citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory~\citep{barabas_chessons_2018, johnson_resolving_2022} is biologically appealing, it is likely far from trivial and may present a promising avenue for future work.

>>>>

At first thought, the idea of an evolutionary force that selects individuals with lower birth and death rates over individuals with higher birth and death rates may be reminiscent of notions in life-history evolution like $r$ vs $K$ selection or selection on the pace of life~\citep{stearns_evolution_1977}. However, it is not possible to conclude whether this similarity reflects some deep principle or whether it is just superficial based solely on the work conducted in this thesis. Models in life-history theory are often primarily concerned with spatiotemporally fluctuating external environments, and thus the stochasticity at the core of life-history models is extrinsic to the population. Ecological frameworks such as modern coexistence theory, which also deal with questions about similar population dynamics and would benefit from a first principles stochastic birth-death formulation, also typically work with fluctuating external environments~\citep{chesson_stabilizing_1982,chesson_multispecies_1994,barabas_chessons_2018, johnson_resolving_2022}. I have entirely neglected such extrinsic factors in my formalism. In principle, it is possible to make the birth and death rates \eqref{nD_functional_forms_for_replicator} in my framework also depend on a temporally varying external environment $\mathbf{E}(t)$ (whose variation may possibly depend on the population $\mathbf{n}(t)$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. If the variation of the environment $\mathbf{E}(t)$ has some associated stochasticity, the complete dynamics of the system would be the result of interactions between two qualitatively different forms of noise, \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population. Both theoretical~\citep{gokhale_eco-evolutionary_2016,wienand_evolution_2017} and empirical~\citep{chavhan_larger_2020, chavhan_interplay_2021} work suggests that such dynamics can be quite complex and intricate, and may consequently be difficult to handle analytically in the sort of generality I have employed throughout this thesis. Thus, while integrating the birth-death framework I outline here with ecological ideas such as the pace-of-life syndrome~\citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory~\citep{barabas_chessons_2018, johnson_resolving_2022} is biologically appealing, it is likely far from trivial and may present a promising avenue for future work.
